Variations on a Bayesian Theme: Comparing Bayesian Models of Referential Reasoning

### 2013-10-30 01_intro.tex
rephrased intro, amended layout

Human language communication is efficient, in that people need not always say every detail in order to be understood. Rather, speakers only say what is most relevant, and listeners can seemingly effortlessly grasp the intended meaning beyond what is literally said. This ability to do pragmatic inference has been long studied and the \emph{conversational implicature} theory by Grice \cite{Grice} is one of the most prominent theories in the field. However, since it is hard to formalize the \emph{Cooperative Principle} and the \emph{Conversational Maxims}, Grice's theory is not precise enough to give empirically testable predictions, especially quantitative predictions.

The Bayesian \emph{Rational Speech Act} (RSA) model attempts to address this issue by using information-theoretic notions together with the Bayesian inference framework. It has been shown that the model could yield quantitative predictions that are highly correlated to actual human judgments \cite{Frank}.

Despite the RSA model's theoretical advantage in that it provides us with quantitative predictions and its empirical success, if we analyze the design of the model, we will find several choices that are not \emph{prima facie} obvious and thus need further clarification. These choices have their alternatives in the closely related game-theoretic approaches to pragmatics (e.g. \cite{Benz2007,Jager2013}), which have similar but slightly different conceptual motivations. Hence it is important to systematically compare the original RSA model with all these alternatives and test their predictions on empirical data, to gain a better understanding of the nature of human pragmatic inference.

The outline of the paper is as follows. We first introduce the simple referential communication game investigated in \cite{Frank} and the original RSA model proposed there. Next we analyze a few design choices in this model and introduce their alternatives in game-theoretic pragmatics. Finally we report on the result of an experiment similar to the original one, and compare the predictions of different models to the new data.

>>>>

Human language communication is efficient, in that people need not
always say every detail in order to be understood. Rather, speakers
often only say what is most relevant, and listeners can often
seemingly effortlessly grasp the intended meaning beyond what is
literally said. This ability to do pragmatic inference has been long
studied and the \emph{conversational implicature} theory by Grice
\cite{Grice} is one of the most prominent theories in the
field. However, since it is hard to formalize the \emph{Cooperative
  Principle} and the \emph{Conversational Maxims}, Grice's theory is
not precise enough to give empirically testable predictions,
especially quantitative predictions.

The Bayesian \emph{Rational Speech Act} (RSA) model attempts to
address this issue by using information-theoretic notions together
with the Bayesian inference framework
\cite{Frank2009,Frank,Bergen2012,GoodmanStuhlmuller2013:Knowledge-and-I}. It
has been shown that the model could yield quantitative predictions
that are highly correlated to actual human judgments
\cite{Frank,GoodmanStuhlmuller2013:Knowledge-and-I}.

Despite the RSA model's theoretical promise and empirical success, if
we analyze the design of the model, we will find several choices that
are not \emph{prima facie} obvious and might thus need further
consideration and justification. These choices have their alternatives
in closely related game-theoretic approaches to pragmatics
(e.g. \cite{Benz2007,Jager2013}), which have similar but slightly
different conceptual motivations. Hence it is important to
systematically compare the original RSA model with all these
alternatives and test their predictions on empirical data, so as to
gain a better understanding of the relation between these
models. Doing so will help illuminate the theoretical commitments and
empirical implications of each design choice, which will enhance our
understanding of the nature of human pragmatic inference.

The outline of the paper is as
follows. Section~\ref{sec:refer-games-rsa} introduce the referential
communication games investigated by \cite{Frank} and the original RSA
model proposed there. Section~\ref{sec:alternatives} analyzes a few
design choices in this model and introduces their alternatives in game
theoretic pragmatics, emphasizing the distinction between
\emph{belief}, \emph{goal} and \emph{action} and unifying various
models as different interpretations of these three
components. Section~\ref{sec:experiment} reports on the results of an
experiment similar to the one of \cite{Frank}. In
Section~\ref{sec:model-comparison}, we compare the predictions of
different models of our new data. Finally, we discuss the implications
of the results of the model comparison in
Section~\ref{sec:conclusion}, concluding that the original RSA model
can be improved on, in particular by taking speaker preferences into
account, but we also point out that a more complex picture of
probabilistic pragmatic reasoning about referential expressions awaits
further exploration.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% TeX-PDF-mode: t
%%% End:

### 2013-10-30 main.tex
rephrased intro, amended layout

\title{Variations on a Bayesian Theme: Comparing Bayesian Models of Referential Reasoning}
\author{Michael Franke \and Ciyang Qing}
\institute{Institute for Logic, Language and Computation}

@@

\begin{abstract}
Recent developments in Bayesian experimental pragmatics have received much attention. The Rational Speech Act (RSA) model formalizes core concepts in traditional pragmatic theories quantitatively and makes predictions that fit empirical data nicely. In this paper, we analyze the RSA model and its relation to the closely related game-theoretic approaches to pragmatics, by spelling out its \emph{belief}, \emph{goal} and \emph{action} components. We introduce some alternatives motivated from the game-theoretic models and compare them with the RSA model to new experimental data. The result suggests that the RSA model could be adapted and extended to improve its predictive power.

\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Referential Games and the RSA model}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Alternatives in Design Choices}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Experiment}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Model Comparison}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}

>>>>

\title{Variations on a Bayesian Theme: Comparing Bayesian Models of
  Referential Reasoning}
\titlerunning{Variation on a Bayesian Theme}
\author{Ciyang Qing \and Michael Franke\thanks{Thanks to \dots % Judith Degen,
  % Noah Goodman, Daniel Lassiter [?], audience of the workshop, Henk
  % & Hans-Christian for organizing, reviewers \dots 
}}
\institute{Institute for Logic, Language and Computation\\Universtity
  of Amsterdam, The Netherlands}

@@

\begin{abstract}
  Recent developments in Bayesian experimental pragmatics have
  received much attention. The Rational Speech Act (RSA) model
  formalizes core concepts of traditional pragmatic theories
  quantitatively and makes predictions that fit empirical data
  nicely. In this paper, we analyze the RSA model and its relation to
  closely related game theoretic approaches, by spelling out its
  \emph{belief}, \emph{goal} and \emph{action} components. We
  introduce some alternatives motivated from the game theoretic
  tradition and compare models incorporating these alternatives
  systematically to the original RSA model, using Bayesian model
  comparison, in terms of their ability to predict relevant empirical
  data. The result suggests that the RSA model could be adapted and
  extended to improve its predictive power, in particular by taking
  speaker preferences into account. 
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Referential Games and the RSA model}
\label{sec:refer-games-rsa}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Alternatives in Design Choices}
\label{sec:alternatives}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Experiment}
\label{sec:experiment}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Model Comparison}
\label{sec:model-comparison}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
\label{sec:conclusion}
